Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-03-30 tagless/tagless-final.tex
A bit more work on introduction

\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}

@@

\usepackage[compact]{fancyvrb1}
\DefineShortVerb{\|}
\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=1em,fontsize=\small}

@@

\newcommand{\fun}[1]{\mathopen{\lambda\mathord{#1}.\,}}
\newcommand{\fix}[1]{\mathopen{\mathrm{fix\,}\mathord{#1}.\,}}
\newcommand{\Forall}[1]{\mathopen{\forall\mathord{#1}.\,}}
\newcommand{\Exists}[1]{\mathopen{\exists\mathord{#1}.\,}}
\newcommand{\cond}[3]{\mathrm{if\,}#1\mathrm{\,then\,}#2\mathrm{\,else\,}#3}

@@

A popular way to define and implement a programming language is to embed it in
another \citep{reynolds-definitional}.  The embedded language is called the
\emph{object} or \emph{subject} language, whereas the host language is the
\emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
specific object languages because it supports rapid prototyping and integration
with the host environment \oleg{Cite Hudaks' survey and perhaps say a
  couple of more words}.  If the metalanguage supports \emph{staging}, then
the embedding can compile object programs to the metalanguage and avoid the
overhead of interpreting them on the fly \citep{WalidICFP02}.  A staged
definitional interpreter is thus a promising way to build a domain\hyp specific
language.

\begin{figure}
    \begin{floatrule}
    \begin{proofrules}
        \[ \[ [x:t_1] \proofoverdots e:t_2 \] \justifies \fun{x}e:t_1\to t_2 \]
        \[ \[ [f:t_1\to t_2] \proofoverdots e:t_1\to t_2 \] \justifies \fix{f}e:t_1\to t_2 \]
        \[ e_1:t_1\to t_2 \quad e_2:t_1 \justifies e_1 e_2: t_2 \]
        \[ \text{$n$ is an integer} \justifies n:\ZZ \]
        \[ \text{$b$ is a Boolean} \justifies b:\BB \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1+e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \times e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \le e_2:\BB \]
        \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
    \end{proofrules}
    \end{floatrule}
    \caption{Our typed object language}
    \label{fig:object}
\end{figure}

We focus on embedding a typed object language into a typed metalanguage.
\ccshan{Collect reasons to type here.}

    At POPL 2003, Taha et al~\cite{TahaPOPL03} and Xi et al~\cite{XiPOPL03}
both
interpreter of a typed language in a typed language is an interesting
problem. The known solutions include the use of the Universal type, or
GADTs. The former is not quite satisfactory for performance reasons;
and also because of the presence of partial pattern matching in the
interpreter; the inexhaustiveness of the pattern matching should never
be triggered when executing a well-typed term. Alas, this fact cannot
be made obvious to the compiler. That also exlains unoptimality.

\oleg{should say somewhere in the Introduction that we are interested
  in the typed object language embedded in the typed
  meta-language. Refer to Sec 2 (see the remark below) why this is
  good. We assume the DSL setting: we enter EDSL terms in the
  meta-language program (e.g., GHCi or (Meta)OCaml top-level). A
  different approach is reading EDSL terms from a file. In that case, we
  need to typecheck them first. We do not deal with that problem here,
  because it has been solved. First we mention Walid02 where they use
  dependent typed. However, these don't seem to be needed: refer to
  `typing dependent typing', and the related work section.}
\jacques{I think it is important to emphasize, probably later in the 
paper, that we could write a camlp4 extension (or simply a parser in
Haskell) that would transform a nice syntax for a lambda calculus into
syntactic terms for what we present here.  Even the self-interpreter
terms could be done this way, which would simplify things tremendously.}
\jacques{I quite agree that the point that needs to be revisited is
  that a \emph{direct} interpreter for a (typed) language "cannot" be
  done. see discussion in related work about the direct interpreter,
  with regards to Walid's paper. Perhaps bring some of it here?}
\jacques{I have looked at a fair bit of the literature, and I have not
  seen this done, especially not in the typed 'online' setting.  The
  one question that seems to be open is, how accurate can we make the
  PE without a tough BTA?}
Should we mention some of these points in our contributions subsection
below?

@@

Consider a language with abstraction, application and boolean
literals with de Bruin indices as variables. This is the language used
by \citet{WalidICFP02}, from which we borrow the first part of 
this example.

The language can be described by the following grammar (using OCaml
syntax)

\begin{code}
  type var = VZ | VS of var
  type exp = V of var | B of bool | A of exp * exp | L of exp
\end{code}

with a sample term (application of the identity function to a boolean) as
\begin{code}
  let test1 = A ((L (V VZ)),(B true))
\end{code}

Let us attempt to implement an the interpreter, the function |eval0|. It takes
an object language term such as |test1| and gives us its value.
The first argument to |eval| is the environment, initially empty,
which is the sequence of values associated with free variables in the
interpreted code.
\begin{code}
  let rec eval0 env = function 
  | V VZ      -> List.hd env
  | V (VS v)  -> eval0 (List.tl env) (V v)
  | B b       -> b 
  | L e       -> fun x -> eval0 (x::env) e
  | A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

Suppose our meta-language (the language in which we wrote |test1|
and that interpreter) were untyped. The above code would be acceptable.
The |L e| line exhibits interpretive overhead \cite{WalidICFP02}: 
|eval0 env' e| will be executed every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead (sections 1.1 and 1.2 of \cite{WalidICFP02} explain
this well).

@@

@@

>>>>

\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}

@@

\usepackage[compact]{fancyvrb1}
\DefineShortVerb{\|}
\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\parindent}

@@

\newcommand{\fun}[1]{\mathopen{\lambda\mathord{#1}.\,}}
\newcommand{\fix}[1]{\mathopen{\mathrm{fix\,}\mathord{#1}.\,}}
\newcommand{\Forall}[1]{\mathopen{\forall\mathord{#1}.\,}}
\newcommand{\Exists}[1]{\mathopen{\exists\mathord{#1}.\,}}
\newcommand{\cond}[3]{\mathrm{if\ }#1\mathrm{\ then\ }#2\mathrm{\ else\ }#3}
\newcommand{\True}{\mathinner{\mathrm{true}}}
\newcommand{\False}{\mathinner{\mathrm{false}}}

@@

A popular way to define and implement a programming language is to embed it in
another \citep{reynolds-definitional}.  Embedding means to represent
terms and values of the \emph{object language} as terms and values in the
\emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
specific object languages because it supports rapid prototyping and integration
with the host environment \citep{hudak-building} \oleg{perhaps say a
  couple more words}.  If the metalanguage supports \emph{staging}, then
the embedding can compile object programs to the metalanguage and avoid the
overhead of interpreting them on the fly \citep{WalidICFP02}.  A staged
definitional interpreter is thus a promising way to build a domain\hyp specific
language.

\begin{figure}
    \begin{floatrule}
    \begin{proofrules}
        \[ \[ [x:t_1] \proofoverdots e:t_2 \] \justifies \fun{x}e:t_1\to t_2 \]
        \[ \[ [f:t_1\to t_2] \proofoverdots e:t_1\to t_2 \] \justifies \fix{f}e:t_1\to t_2 \]
        \[ e_1:t_1\to t_2 \quad e_2:t_1 \justifies e_1 e_2: t_2 \]
        \[ \text{$n$ is an integer} \justifies n:\ZZ \]
        \[ \text{$b$ is a Boolean} \justifies b:\BB \]
        \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1+e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \times e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \le e_2:\BB \]
    \end{proofrules}
    \end{floatrule}
    \caption{Our typed object language}
    \label{fig:object}
\end{figure}

We focus on embedding a typed object language into a typed metalanguage.
Types are useful as usual because they document the programmer's
intentions and can be checked early by machine.  \ccshan{Collect reasons
to type here.}  To be concrete, we use the typed object language in
Figure~\ref{fig:object} throughout this paper.  We aim not just to
interpret the object language in the narrow sense but also for
compilation, partial evaluation, and other processing.

\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
a typed object language in a typed metalanguage as an interesting
problem.  The known solutions to this problem store object terms and
values in the metalanguage in a universal type, a generalized algebraic
datatype (GADT), or a dependent type.  In the remainder of this section,
we discuss these solutions, identify their drawbacks, then summarize our
contributions.  No matter how we represent the object language in the
metalanguage, the representation can be performed by hand or more
conveniently by a parser or preprocessor such as camlp4.  We leave aside
the issue of whether to program a parser or preprocessor using dependent
types \citep{WalidICFP02} or not \citep{baars-typing}.

@@

It is straightforward to create an algebraic data type, say in OCaml, to
represent object terms such as those in Figure~\ref{fig:object}.
\begin{code}
type var = VZ | VS of var
type exp = V of var | B of bool
         | L of exp | A of exp * exp | ...
\end{code}
We represent each variable using a unary de-Bruijn number.
For example, we represent the object term $(\fun{x}x)\True$ as
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}

Following \citet{WalidICFP02},
we try to implement an interpreter function |eval0|. It takes
an object term such as |test1| above and gives us its value.
The first argument to |eval0| is the environment, initially empty,
which is the list of values bound to free variables in the
interpreted code.
\begin{code}
let rec lookup (x::env) = function
| VZ   -> x
| VS v -> lookup env v

let rec eval0 env = function 
| V v       -> lookup env v
| B b       -> b 
| L e       -> fun x -> eval0 (x::env) e
| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}
If our OCaml-like metalanguage were untyped, the code above would be acceptable.
The |L e| line exhibits interpretive overhead:
|eval0| traverses the function body~|e| every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead \cite[\S1.1--2]{WalidICFP02}.

@@

The universal type fails to express in the metalanguage
that the object program is well-typed.  This failure gives rise to
tagging and untagging (that is, apparently inexhaustive pattern\hyp
matching) operations that incur a performance penalty in interpretation
and impair optimality in partial evaluation.  In short, the
universal\hyp type solution is not type\hyp preserving.

@@

\jacques{I quite agree that the point that needs to be revisited is
  that a \emph{direct} interpreter for a (typed) language "cannot" be
  done. see discussion in related work about the direct interpreter,
  with regards to Walid's paper. Perhaps bring some of it here?}
\jacques{I have looked at a fair bit of the literature, and I have not
  seen this done, especially not in the typed 'online' setting.  The
  one question that seems to be open is, how accurate can we make the
  PE without a tough BTA?}
Should we mention some of these points in our contributions subsection
below?
